Mental Rotation as Bayesian Quadrature

### 2013-10-16 man/nips-2013-bayesian-optimization.tex
Small edits, mostly rewording

\begin{abstract}
  Given a computational resource--for example, the ability to
  visualize an object rotating--how do you best make use of it? We
  explored how mental simulation should used in the classic
  psychological task of determining if two images depict the same
  object in different orientations. We compared two models on this
  mental rotation task, and found that a model based on an optimal
  experiment design for Bayesian quadrature is objectively more
  accurate and qualitatively more consistent with classic behavioral
  data than a simpler model. We suggest that rational models which
  adaptively exploit available resources are promising in their
  ability to characterize metacognitive processes like mental
  simulation.
\end{abstract}

@@

We begin by first analyzing mental rotation at Marr's
\textit{computational} level \cite{Marr:1983to}: what is the problem
to be solved?  Formally, people are presented with two images, $X_a$
and $X_b$, which are the coordinates of the vertices of 2D shapes
(e.g., Fig.~\ref{fig:stimuli}). Participants must determine whether
$X_a$ and $X_b$ depict the same shape, i.e., whether $\exists
R\textrm{ s.t. } X_b=RX_a$, where $R$ is a rotation matrix. We can
formulate the judgment of whether $X_a$ and $X_b$ have the same
origins by deciding about two hypotheses, $h_0$: $\forall R\ X_b\neq
RX_a$ and $h_1$: $\exists R\textrm{ s.t. } X_b=RX_a$.  To compare the
hypotheses, we need to compute the posterior for each: $p(h\ \vert\
X_a, X_b)\propto p(X_a, X_b\ \vert\ h)p(h)$. Assuming the hypotheses
are equally likely \textit{a priori}, the prior term $p(h)$ will
cancel out when comparing $h_0$ and $h_1$, thus allowing us to focus
on the likelihoods, which are $p(X_a,\ X_b\ \vert \ h_0)=p(X_a)p(X_b)$
for $h_0$ and $p(X_a,\ X_b\ \vert \ h_1)=\int_R p(X_a) p(X_b\vert
X_a,R) p(R)\ \mathrm{d}R$ for $h_1$. From these likelihoods, we
compute the ratio $\ell$ which is given by $\ell=\left(\int_R p(X_b\
  \vert\ X_a, R)p(R)\ \mathrm{d}R\right) /\ p(X_b)$. If $\ell<1$, then
$h_0$ is the more likely hypothesis. If $\ell>1$, then $h_1$ is the
more likely hypothesis.

@@

A more flexible strategy uses what is known as \emph{Bayesian
  Quadrature} (BQ) \cite{Diaconis:1988uo,OHagan:1991tx} to estimate
$Z$.  BQ computes a posterior distribution over $Z$ by placing a
Gaussian Process (GP) prior on the function $S$ and evaluating $S$ at
a particular set of points. However, althought $S$ is a non-negative
likelihood function, GP regression enforces no such
constraint. \cite{Osborne:2012tm} give a method to place a prior over
the log likelihood,\footnote{In practice, as in \cite{Osborne:2012tm},
  we use the transform of $\log(S+1)$. Additionally, we follow
  \cite{Osborne:2012tm} and use a combination of MLII and
  marginalization to fit the $w$ kernel parameters. The output scales
  are fixed at $h_S=\sqrt{0.1}$ and $h_{\log S}=\log(h_S + 1)$.} thus
ensuring that $S=e^{\log S}$ will be positive, i.e.  $E[Z\ \vert \
\log S]=\int_{\log S}\left(\int_R \exp(\log{S(X_b,X_R)})p(R)\
  \mathrm{d}R\right)\mathcal{N}\left(\log{S}\ \vert \ \mu_{\log S},
  \Sigma_{\log S}\right)\ \mathrm{d}\log S$, where $\mu_{\log S}$ and
$\Sigma_{\log S}$ are the mean and covariance, respectively, of the GP
over $\log S$ given $\mathbf{R}$. We approximate this according to the
method given in \cite{Osborne:2012tm}, i.e. $\mu_Z=E[Z\ \vert \ S,
\log S, \Delta_c] \approx \int_R \mu_{S}(1 + \mu_{\Delta_c}) p(R)\
\mathrm{d}R$, where $\mu_S$ is the mean of a GP over $S$ given
$\mathbf{R}$; and $\mu_{\Delta_c}$ of a GP over $\Delta_c=\mu_{\log S}
- \log \mu_S$ given $\mathbf{R}_c$, which consists of $\mathbf{R}$ and
a set of intermediate \emph{candidate points} $c$ as described in
\cite{Osborne:2012tm}. The variance is $\tilde{V}(Z\vert S, \log S,
\Delta_c)$ as defined in Equation 12 of \cite{Osborne:2012tm}.

@@

We pick an initial direction of rotation which results in the higher
value of $S$. At each step, we compute $\mu_Z$ and $\tilde{V}$ to
estimate a distribution over the likelihood ratio $\ell$, i.e.
$p(\ell)\approx\frac{1}{p(X_b)}\ \mathcal{N}(Z\ \vert\ \mu_Z,
\sigma_z)$.  We choose $h_0$ when $p(\ell < 1)\geq 0.95$, and $h_1$
when $p(\ell > 1)\geq 0.95$. Until one of these conditions are met (or
the shape has been fully rotated), the model will continue to compute
rotations and update its estimate of $Z$.  The model may also change
direction based on the \textit{expected} posterior variance of $Z$
given some new sample $a$.\footnote{This is similar to the full
  posterior variance calculation given in \cite{Osborne:2012tm},
  however we compute the variance given only the current mean estimate
  of $a$.} The model will change directions if doing so would lower
this expected variance. Thus, it is able to actively change its
strategy, unlike the hill-climbing procedure.

@@

We evaluated each model's performance on 20 randomly generated shapes
which had between three and six vertices, inclusive, (e.g., $X_a$ in
Fig.~\ref{fig:stimuli}). For each shape, we computed 18 ``same'' and
18 ``different'' stimuli pairs, with $R$ spaced at $20^\prime$
increments between 0 and 360, as in \cite{Shepard1971}. ``Same'' pairs
were created by rotating $X_a$ by $R$; the same was true for
``different'', except that $X_a$ was also reflected across the
$y$-axis. To gauge performance, we looked at \textit{response error
  rates}: how accurate was the model at choosing the correct
hypothesis? This was defined as the mean error ($\ME{}$), or fraction
of times the model chose incorrectly.  We additionally looked at
\textit{rotations}: for those ``same'' pairs which the model judged
correct, how correlated were the model's rotations with the true
angles of rotation?  We quantified this using the Pearson's
correlation coefficient $\rho$ for the true rotation ($R$) versus the
number of steps/rotations take by the model ($\vert
\mathbf{R}\vert$). Fig.~\ref{fig:rotations} shows true rotations vs
number of steps/rotations by the model for individual stimuli
(\ref{fig:all-stimuli}) and average per true rotation
(\ref{fig:correct-stimuli}). This latter analysis can be qualitatively
compared to the results of \cite{Shepard1971}
(Fig.~\ref{fig:mental-rotation}).

@@

The \naive{} model's response error rate was $\NaiveME{}$, which is
better than chance (equivalent to guessing randomly,
i.e. $\ME{}=0.5$). The correlation between the \naive{} model's
average rotation and the true angle of rotation was $\Naivecorr{}$
(Fig.~\ref{fig:correct-stimuli}, left). For $R<\frac{\pi}{2}$, the
\naive{} model corresponds extremely well to the true angle of
rotation; this is because it needs to rotate less and is therefore
less likely to get stuck on local maxima. For $R>\frac{\pi}{2}$, we
see an increasing tendency to under-rotate due to getting stuck on
local maxima, as well as a tendency to over-rotate if the wrong
direction was initially chosen.

@@

The BQ model was much more accurate in choosing the correct hypothesis
than the other model ($\BQME{}$). The number of rotations computed by
the BQ model were strongly correlated with the true rotations
($\BQcorr$), a result which is qualitatively similar to that exhibited
by humans (Fig.~\ref{fig:mental-rotation}
vs. Fig.~\ref{fig:correct-stimuli}, right). Because the BQ model has
the capacity to ``reset'', it could recover from rotating in the
incorrect direction (e.g., Fig.~\ref{fig:bq}) and thus did not
over-rotate as frequently.  It also under-rotated less frequently:
because the model continues to rotate until it is confident in its
estimate of $Z$, it does not get stuck as easily on local optima.

>>>>

\begin{abstract}
  Given a computational resource--for example, the ability to
  visualize an object rotating--how do you best make use of it? We
  explored how mental simulation should be used in the classic
  psychological task of determining if two images depict the same
  object in different orientations. We compared two models on this
  mental rotation task, and found that a model based on an optimal
  experiment design for Bayesian quadrature is objectively more
  accurate and qualitatively more consistent with classic behavioral
  data than a simpler model. We suggest that rational models which
  adaptively exploit available resources are promising in their
  ability to characterize metacognitive processes like mental
  simulation.
\end{abstract}

@@

We begin by analyzing mental rotation at Marr's \textit{computational}
level \cite{Marr:1983to}: what is the problem to be solved?  Formally,
people are presented with two images, $X_a$ and $X_b$, which are the
coordinates of the vertices of 2D shapes (e.g.,
Fig.~\ref{fig:stimuli}). Participants must determine whether $X_a$ and
$X_b$ depict the same shape, i.e., whether $\exists R\textrm{ s.t. }
X_b=RX_a$, where $R$ is a rotation matrix. We can formulate the
judgment of whether $X_a$ and $X_b$ have the same origins by deciding
about two hypotheses, $h_0$: $\forall R\ X_b\neq RX_a$ and $h_1$:
$\exists R\textrm{ s.t. } X_b=RX_a$.  To compare the hypotheses, we
need to compute the posterior for each: $p(h\ \vert\ X_a, X_b)\propto
p(X_a, X_b\ \vert\ h)p(h)$. Assuming the hypotheses are equally likely
\textit{a priori}, the prior term $p(h)$ will cancel out when
comparing $h_0$ and $h_1$, thus allowing us to focus on the
likelihoods, which are $p(X_a,\ X_b\ \vert \ h_0)=p(X_a)p(X_b)$ for
$h_0$ and $p(X_a,\ X_b\ \vert \ h_1)=\int_R p(X_a) p(X_b\vert X_a,R)
p(R)\ \mathrm{d}R$ for $h_1$. From these likelihoods, we compute the
ratio $\ell$ which is given by $\ell=\left(\int_R p(X_b\ \vert\ X_a,
  R)p(R)\ \mathrm{d}R\right) /\ p(X_b)$. If $\ell<1$, then $h_0$ is
the more likely hypothesis. If $\ell>1$, then $h_1$ is the more likely
hypothesis.

@@

A more flexible strategy uses what is known as \emph{Bayesian
  Quadrature} (BQ) \cite{Diaconis:1988uo,OHagan:1991tx} to estimate
$Z$.  BQ computes a posterior distribution over $Z$ by placing a
Gaussian Process (GP) prior on the function $S$ and evaluating $S$ at
a particular set of points. However, while $S$ is a non-negative
likelihood function, GP regression enforces no such
constraint. \cite{Osborne:2012tm} give a method to place a prior over
the \textit{log} likelihood, thus ensuring that $S=e^{\log S}$ will be
positive, i.e.  $E[Z\ \vert \ \log S]=\int_{\log S}\left(\int_R
  \exp(\log{S(X_b,X_R)})p(R)\
  \mathrm{d}R\right)\mathcal{N}\left(\log{S}\ \vert \ \mu_{\log S},
  \Sigma_{\log S}\right)\ \mathrm{d}\log S$, where $\mu_{\log S}$ and
$\Sigma_{\log S}$ are the mean and covariance, respectively, of the GP
over $\log S$ given $\mathbf{R}$. We approximate this according to the
method given in \cite{Osborne:2012tm}, i.e. $\mu_Z=E[Z\ \vert \ S,
\log S, \Delta_c] \approx \int_R \mu_{S}(1 + \mu_{\Delta_c}) p(R)\
\mathrm{d}R$, where $\mu_S$ is the mean of a GP over $S$ given
$\mathbf{R}$; and $\mu_{\Delta_c}$ of a GP over $\Delta_c=\mu_{\log S}
- \log \mu_S$ given $\mathbf{R}_c$, which consists of $\mathbf{R}$ and
a set of intermediate \emph{candidate points} $c$ as described in
\cite{Osborne:2012tm}. The variance is $\tilde{V}(Z\vert S, \log S,
\Delta_c)$ as defined in Equation 12 of \cite{Osborne:2012tm}.

@@

We pick the initial direction of rotation which results in the higher
value of $S$. From then on, at each step we compute $\mu_Z$ and
$\tilde{V}$ to estimate a distribution over the likelihood ratio
$\ell$, i.e.  $p(\ell)\approx\frac{1}{p(X_b)}\ \mathcal{N}(Z\ \vert\
\mu_Z, \sigma_z)$.  We choose $h_0$ when $p(\ell < 1)\geq 0.95$, and
$h_1$ when $p(\ell > 1)\geq 0.95$. Until one of these conditions are
met (or the shape has been fully rotated), the model will continue to
compute rotations and update its estimate of $Z$.  The model may also
change direction based on the \textit{expected} posterior variance of
$Z$ given some new sample $a$.\footnote{This is similar to the full
  posterior variance calculation given in \cite{Osborne:2012tm},
  however we compute the variance given only the current mean estimate
  of $a$.} The model will change directions if doing so would lower
this expected variance. Thus, it is able to actively change its
strategy, unlike the hill-climbing procedure.

@@

We evaluated each model's performance on 20 randomly generated shapes
which had between three and six vertices, inclusive, (e.g., $X_a$ in
Fig.~\ref{fig:stimuli}). For each shape, we computed 18 ``same'' and
18 ``different'' stimuli pairs, with $R$ spaced at $20^\prime$
increments between 0 and 360, as in \cite{Shepard1971}. ``Same'' pairs
were created by rotating $X_a$ by $R$; the same was true for
``different'' pairs, except that $X_a$ was also reflected across the
$y$-axis. To gauge performance, we looked at \textit{response error
  rates}: how accurate was the model at choosing the correct
hypothesis? This was defined as the mean error ($\ME{}$), or fraction
of times the model chose incorrectly.  We additionally looked at
\textit{rotations}: for those ``same'' pairs which the model judged
correctly, how correlated were the model's rotations with the true
angles of rotation?  We quantified this using the Pearson's
correlation coefficient $\rho$ for the true rotation, $R$, versus the
number of steps/rotations take by the model, $\vert
\mathbf{R}\vert$. Fig.~\ref{fig:rotations} shows true rotations vs
(\ref{fig:all-stimuli}) the number of steps/rotations taken by the
model for individual stimuli, and (\ref{fig:correct-stimuli}) the
average number of steps taken for each true rotation. This latter
analysis can be qualitatively compared to the results of
\cite{Shepard1971}, as shown in Fig.~\ref{fig:mental-rotation}.

@@

The \naive{} model's response error rate was $\NaiveME{}$, which is
better than chance (where chance is equivalent to guessing randomly,
i.e. $\ME{}=0.5$). The correlation between the \naive{} model's
average rotation and the true angle of rotation was $\Naivecorr{}$
(Fig.~\ref{fig:correct-stimuli}, left). As shown in
Fig.~\ref{fig:correct-stimuli} (right), the \naive{} model corresponds
extremely well to the true angle of rotation when $R<\frac{\pi}{2}$;
this is because it needs to rotate less and is therefore less likely
to get stuck on local maxima. For $R>\frac{\pi}{2}$, we see an
increasing tendency to under-rotate due to getting stuck on local
maxima, as well as a tendency to over-rotate if the wrong direction
was initially chosen.

@@

The BQ model was much more accurate in choosing the correct hypothesis
than the \naive{} model ($\BQME{}$). The number of rotations computed
by the BQ model were strongly correlated with the true rotations
($\BQcorr$), a result which is qualitatively similar to that exhibited
by humans (Fig.~\ref{fig:mental-rotation}
vs. Fig.~\ref{fig:correct-stimuli}, right). Because the BQ model has
the capacity to ``reset'', it could recover from rotating in the
incorrect direction (e.g., Fig.~\ref{fig:bq}) and thus did not
over-rotate as frequently as the \naive{} model.  The BQ model also
under-rotated less frequently because it rotates until it is confident
in its estimate of $Z$ and thus does not get stuck on local optima.
